\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2020-12-09 paper.tex
lights v8

\usepackage[colorlinks=true,
            linkcolor=red,
            urlcolor=blue,
            citecolor=blue,
            breaklinks=true]{hyperref}
\usepackage{breakurl} 
\usepackage[a4paper, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
\RequirePackage{amsmath,amsfonts, amssymb,amsthm}

@@

\usepackage{authblk}
\title{\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}}
\author[1,2]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
\author[3]{Antoine Barbieri}
\author[1]{Sarah Zohar}
\author[1,4]{Anne-Sophie Jannot}
\affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
\affil[2]{Califrais' Machine Learning Lab, Paris, France}
\affil[3]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
\affil[4]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{xcolor}
\usepackage{mathrsfs} 
\usepackage{natbib}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{subfigure}

@@

An assumption of heterogeneity within the subject population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}. To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}. Let us denote
\begin{equation}
  \label{eq:pi}
  \pi_{\xi_k}(x) = \P[G=k|X=x]
\end{equation}
the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
\[ \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}} \]
where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero. The intercept term is here omitted without loss of generality. 
From now on, all computations are done conditionally on features $x$.

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independant features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during inference, see~\eqref{eq:pen-log-lik}. Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$. 
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
and
\begin{equation}
  \label{eq:b_density}
  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
\end{equation}

@@

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$ and $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
Note that in the $\gamma_k$ penalty, $l$ actually starts at 0 (and not 1), but for simplicity we do not change the notation. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independant features involved in the $k$-th group membership. The ridge part allows to handle potential correlation between features.
On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
Finally, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory $l$ with the right complexity.
Note that in practice, the intercept is not regularized.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g(b_i, G_i) = \tilde g(b_i)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.

@@

\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

@@

\paragraph*{Initialization.}
Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced. We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:QNMCEM}. 
For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$. Then, we initialize $\lambda_0^{(0)}$ and $\gamma_{k,0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using \texttt{R} package \texttt{survival}. Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm, being itself initialized with univariates fits using \texttt{R} package \texttt{nlme}.

@@

\begin{algorithm}[]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

@@

\paragraph*{Penalized Cox model with time-independant features.}

>>>>

\usepackage{xcolor}
\definecolor{darkpastelgreen}{rgb}{0.01, 0.75, 0.24}

\usepackage[colorlinks=true,
            linkcolor=red,
            urlcolor=darkpastelgreen,
            citecolor=blue,
            breaklinks=true]{hyperref}
\usepackage{breakurl} 
\usepackage[a4paper, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
\RequirePackage{amsmath,amsfonts, amssymb,amsthm}

@@

\usepackage{authblk}
\title{\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}}
\author[1,2]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
\author[2]{Van Tuan Nguyen}
\author[3]{Antoine Barbieri}
\author[1]{Sarah Zohar}
\author[1,4]{Anne-Sophie Jannot}
\affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
\affil[2]{Califrais' Machine Learning Lab, Paris, France}
\affil[3]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
\affil[4]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{mathrsfs} 
\usepackage{natbib}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{caption}
\usepackage{booktabs}
\usepackage{subfigure}

@@

An assumption of heterogeneity within the subject population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}. To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}. Let us denote
\begin{equation}
  \label{eq:pi}
  \pi_{\xi_k}(x) = \P[G=k|X=x]
\end{equation}
the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
\begin{equation}
  \label{eq:pi-softmax} 
  \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}
\end{equation}
where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero. The intercept term is here omitted without loss of generality. 
From now on, all computations are done conditionally on features $x$.

@@

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}. Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$. 
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,L\cA}^\top)= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]

@@

Let us denote 
\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that for ease of notations, we include $\lambda_0$ in $\theta$ although it is a function.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
\begin{assumption}
\label{indep-hyp-3}
Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
  \label{eq:ind-hyp}
  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$.
\end{assumption}
Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
and
\begin{equation}
  \label{eq:b_density}
  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
\end{equation}

@@

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_{k,0}}_{\text{en}, \eta}\big) + \zeta_{2,k} \norm{\gamma_{k,L\cA}}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$ and $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independent features involved in the $k$-th group membership. The ridge part allows to handle potential correlation between features.
On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
Finally, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory $l$ with the right complexity.
Note that in practice, the intercept is not regularized.

@@

We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
\begin{equation*}
  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
\end{equation*}
Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
\[ \E_{\theta^{(w)}}[ g_i(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
for different functions $g_i$, where we denote 
\begin{equation}
  \label{eq:pi_ik-def}
  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
\end{equation}
the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
Then, either $g_i(b_i, G_i) = \tilde g_i(b_i)$ (for instance $\tilde g_i: b_i \mapsto b_i^\top D^{-1}b_i$), or $g_i(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.

In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice. Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context. We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result. Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.

@@

\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
\begin{equation}
  \label{eq:def-hat-E-g-tilde}
  \hat \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:hat_Lambda_ik}
  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
\end{equation}

@@

\paragraph*{Initialization.}
Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced. We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:QNMCEM}. 
For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$. Then, we initialize $\lambda_0^{(0)}$ and $\gamma_{k,0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}. Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (the details are given in Appendix~\ref{sec:MLMM}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.

@@

\begin{algorithm}[]
\DontPrintSemicolon
\caption{QNMCEM algorithm for the lights model inference}
\label{alg:QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}

@@

\paragraph*{Penalized Cox model with time-independent features.}

The first model we consider as a baseline is the well known Cox PH model with time-independent features.
In this model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
\begin{equation*}
  \ell_n^{\text{cox}}(\beta) = n^{-1} \sum_{i=1}^n \delta_i \big( x_i^\top \beta - \log \sum_{i' : t_{i'} \geq t_i} \text{exp}(x_{i'}^\top \beta) \big).
\end{equation*}
We use respectively the \texttt{R} packages \texttt{survival} and \texttt{glmnet} \citep{simon2011regularization} for the partial log-likelihood and the minimization of the following quantity
\begin{equation*}
  - \ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big),
\end{equation*}
where $\xi$ is chosen by the a 10-fold cross-validation procedure, for a given $\eta \in [0, 1]$. Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}. We also choose to include basic time-independent features extracted from longitudinal processes, namely the average value for each time-dependant feature.
